---
title: "Google Data Analytics Capstone Case Study"
author: "Frank Amo Agyei-Owusu"
date: "4/25/2023"
output: html_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```


## Loading Packages
Here,the tidyverse, janitor and lubridate package were first installed and then loaded to aid in the data cleaning process


```{r}
library(tidyverse)
library(janitor)
library(lubridate)
```


## Loading the Datasets
Here, the data was from 12 csv files and then loaded

```{r}
Jan2022<-read.csv("Google Data Analytics_ Case Study1\\divvy-tripdata\\2022_01.csv")
Feb2022<-read.csv("Google Data Analytics_ Case Study1\\divvy-tripdata\\2022_02.csv")
Mar2022<-read.csv("Google Data Analytics_ Case Study1\\divvy-tripdata\\2022_03.csv")
Apr2022<-read.csv("Google Data Analytics_ Case Study1\\divvy-tripdata\\2022_04.csv")
May2022<-read.csv("Google Data Analytics_ Case Study1\\divvy-tripdata\\2022_05.csv")
Jun2022<-read.csv("Google Data Analytics_ Case Study1\\divvy-tripdata\\2022_06.csv")
Jul2022<-read.csv("Google Data Analytics_ Case Study1\\divvy-tripdata\\2022_07.csv")
Aug2022<-read.csv("Google Data Analytics_ Case Study1\\divvy-tripdata\\2022_08.csv")
Sep2022<-read.csv("Google Data Analytics_ Case Study1\\divvy-tripdata\\2022_09.csv")
Oct2022<-read.csv("Google Data Analytics_ Case Study1\\divvy-tripdata\\2022_10.csv")
Nov2022<-read.csv("Google Data Analytics_ Case Study1\\divvy-tripdata\\2022_11.csv")
Dec2022<-read.csv("Google Data Analytics_ Case Study1\\divvy-tripdata\\2022_12.csv")

```


# Combining or Merging the datasets
The datasets was for individual months so it was combined to one based on the rows to help us have an individual dataset

```{r}
combined_df<-bind_rows(Jan2022, Feb2022, Mar2022, Apr2022, May2022, Jun2022,
                       Jul2022, Aug2022, Sep2022, Oct2022, Nov2022, Dec2022)
```

## Data Cleaning Process
Here, we combined various data cleaning methods to help make the dataset as organized as possible for the data analysis

# Cleaning the dataset to remove any spaces or parentheses

```{r}
combined_df<- clean_names(combined_df)
```

# Removing empty columns and rows in the dataframe

```{r}
remove_empty(combined_df, which = c())
```


## Adding newly columns specified in the Business Task
Here, new columns would be added to the dataset as specified or demanded from the task given in the case study


# day of the week
We use the wday() function here-- it extracts the DAY from the OBJECT(usually SELECTED from a column in a dataframe) with the date format

```{r}
combined_df$day_of_week<-wday(combined_df$started_at, label = T, abbr = T) 
```


# start hour
We use the POSIXct() function-- it extracts a certain TIME HOUR FORMAT from an OBJECT (which is usually formatted as a date/time format)

```{r}
combined_df$starting_hour<-format(as.POSIXct(combined_df$started_at), '%H')
```


# Month
The as.Date() function was deployed to extract the DATE from the OBJECT (usually from a column in a dataframe) with the date format. Used in format()


```{r}
combined_df$month< format(as.Date(combined_df$started_at), '%m')
```


# Trip Duration
The difftime() function was deployed to calculate the time difference between column and another within a dataframe with a date format

```{r}
combined_df$trip_duration<- difftime(combined_df$ended_at, combined_df$started_at, units = 'sec')
```



# Cleaning the data again to remove the errors in the new columns added
The data is then cleaned again to remove any negative trip_durations

```{r}
cleaned_df<-combined_df[!(combined_df$trip_duration <=0),]
```


# Saving the data as a csv file 
The data was saved as a csv to aid in data visualization

```{r}
write.csv(cleaned_df, file = "Cyclist_trip_data.csv")
```





